\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-23 papers/jfp/3-abstractions.tex
Add another todo to AM
Work on S3

\section{Build systems, abstractly}\label{sec-abstractions}
\vspace{-0.5mm}

\begin{figure}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store containing a key/value map and persistent build information
data Store i k v -- i = info, k = key, v = value
initialise :: i -> (k -> v) -> Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
\end{minted}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
data Hash v -- a compact summary of a value with a fast equality check
hash    :: Hashable v => v -> Hash v
getHash :: Hashable v => k -> Store i k v -> Hash v
\end{minted}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build tasks (see §3.2)
newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system (see §3.3)
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system components: a scheduler and a rebuilder (see §5)
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
\vspace{-3.5mm}
\caption{Type signatures of key build systems abstractions.}\label{fig-types}
\vspace{-4.5mm}
\end{figure}

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed in~\S\ref{sec-background},
and design complex build systems from simple primitives. Specifically, we
present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-build}
and~\S\ref{sec-implementations} scrutinise the abstractions further and provide
concrete implementations for several build systems.

\vspace{-1.5mm}
\subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
\vspace{-0.5mm}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is the identity of the monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation of the monoid m
\end{minted}
\vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
\vspace{-4mm}
\end{figure}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: Eq k => Build Applicative () k v
busy tasks key store = execState (fetch key) store
  where
    fetch :: k -> State (Store () k v) v
    fetch k = case tasks k of
        Nothing   -> gets (getValue k)
        Just task -> do v <- run task fetch; modify (putValue k v); return v
\end{minted}
\vspace{1mm}

@@

\subsection{The Need for Polymorphism in Task}\label{sec-why-polymorphism}
\vspace{-0.5mm}

@@

\vspace{-0.5mm}
\subsection{Monadic Tasks}\label{sec-task-monad}

@@

\label{def-correct}
\vspace{2mm}

@@

\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 1
B2: 10
(10,[("C1",1),("B2",10)])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 2
A2: 20
(20,[("C1",2),("A2",20)])
\end{minted}
\vspace{1mm}

@@

>>>>

\section{Build systems, abstractly}\label{sec-abstractions}
% \vspace{-0.5mm}

\begin{figure}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store containing a key/value map and persistent build information
data Store i k v -- i = info, k = key, v = value
initialise :: i -> (k -> v) -> Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
data Hash v -- a compact summary of a value with a fast equality check
hash    :: Hashable v => v -> Hash v
getHash :: Hashable v => k -> Store i k v -> Hash v
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build tasks (see §3.2)
newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system (see §3.3)
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
% \vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system components: a scheduler (see §4) and a rebuilder (see §5)
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
% \vspace{-3.5mm}
\caption{Type signatures of key build systems abstractions.}\label{fig-types}
% \vspace{-4.5mm}
\end{figure}

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed in~\S\ref{sec-background},
and design complex build systems from simple primitives. Specifically, we
present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-scheduler},
\S\ref{sec-rebuilder} and~\S\ref{sec-implementations} scrutinise the
abstractions further and provide concrete implementations for several build
systems.

% \vspace{-1.5mm}
\subsection{Common Vocabulary for Build Systems}\label{sec-vocabulary}
% \vspace{-0.5mm}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is the identity of the monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation of the monoid m
\end{minted}
% \vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
% \vspace{-4mm}
\end{figure}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: Eq k => Build Applicative () k v
busy tasks key store = execState (fetch key) store
  where
    fetch :: k -> State (Store () k v) v
    fetch k = case tasks k of
        Nothing   -> gets (getValue k)
        Just task -> do v <- run task fetch
                        modify (putValue k v)
                        return v
\end{minted}
\vspace{1mm}

@@

\subsection{The Need for Polymorphism in Task}\label{sec-why-polymorphism}
% \vspace{-0.5mm}

@@

% \vspace{-0.5mm}
\subsection{Monadic Tasks}\label{sec-task-monad}

@@

\label{def-correct}
% \vspace{2mm}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 1
B2: 10
(10,[("C1",1),("B2",10)])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 2
A2: 20
(20,[("C1",2),("A2",20)])
\end{minted}
\vspace{1mm}

@@

\subsection{Examples of Tasks}\label{sub-task-examples}

The \hs{Task} abstraction can be used to express pure functions in a way that is
convenient for their memoization, as will be further discussed
in~\S\ref{sec-related-memo}. Below we give examples of tasks corresponding to
several well-known functions.

The \emph{Fibonacci sequence} $F_n = F_{n-1} + F_{n-2}$ is the one of the most
favourite examples in the functional programming community and so we start with
it:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
fibonacci :: Tasks Applicative Integer Integer
fibonacci n = if n < 2 then Nothing else
    Just $ Task $ \fetch -> (+) <$> fetch (n - 1) <*> fetch (n - 2)
\end{minted}
\vspace{1mm}

\noindent
Here the keys $n < 2$ are input parameters, and one can obtain the usual
Fibonacci sequence by picking 0 and 1 for $n = 0$ and $n = 1$, respectively.
Any minimal build system will compute the sequence with memoization, i.e.
without recomputing the same value twice.

% TODO: Add a figure with a dependency graph?
The dependencies of Fibonacci numbers are statically known, hence we can express
it using \hs{Tasks}~\hs{Applicative}, benefiting from static dependency
analysis:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies (fromJust $ fibonacci 5)
[4,3]
\end{minted}
\vspace{1mm}

\noindent
Interestingly, the \emph{Ackermann function} -- a famous example of a function
that is not primitive recursive -- cannot be expressed as an applicative task,
since it needs to perform an intermediate recursive call to determine one of its
dependencies $A(m, n-1)$:

\[
A(m, n) = \left\{
\begin{array}{ll}
n+1 & \mbox{if } m = 0 \\
A(m-1, 1) & \mbox{if } m > 0 \mbox{ and } n = 0 \\
A(m-1, A(m, n-1)) & \mbox{if } m > 0 \mbox{ and } n > 0.
\end{array}\right.
\]

We therefore use \hs{Tasks}~\hs{Monad} to express this function, binding the
dynamic dependency to variable \hs{index}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
ackermann :: Tasks Monad (Integer, Integer) Integer
ackermann (n, m)
    | m < 0 || n < 0 = Nothing
    | m == 0    = Just $ Task $ const $ pure (n + 1)
    | n == 0    = Just $ Task $ \fetch -> fetch (m - 1, 1)
    | otherwise = Just $ Task $ \fetch -> do index <- fetch (m, n - 1)
                                             fetch (m - 1, index)
\end{minted}
\vspace{1mm}

\noindent
Now that we have seen examples of applicative and monadic tasks, let us finish
with an example of a functorial task -- the \emph{Collatz sequence}, where
given an initial value $c_0$, we calculate the next value $c_n$ from $c_{n-1}$
either by dividing $c_{n-1}$ by 2 (if it is even) or multiplying it by 3 and
adding 1 (if it is odd):

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
collatz :: Tasks Functor Integer Integer
collatz n | n <= 0    = Nothing
          | otherwise = Just $ Task $ \fetch -> f <$> fetch (n - 1)
  where
    f k | even k    = k `div` 2
        | otherwise = 3 * k + 1
\end{minted}
\vspace{1mm}

\noindent
Functorial tasks correspond to computations with a linear dependency chain. For
example, computing the element $c_8$ of the Collatz sequence starting from
$c_0=6$ leads to the following dependency chain:
$c_0 = 6 \rightarrow 3 \rightarrow 10 \rightarrow 5 \rightarrow 16 \rightarrow 8
\rightarrow 4 \rightarrow 2 \rightarrow 1 = c_8$.

Collatz sequence is a good example of the early cutoff optimisation
(\S\ref{sec-background-shake}): if we recompute $c_8$ starting from a different
initial value $c_0=40$, the resulting computation will have a large overlap with
the previous one:
$c_0 = 40 \rightarrow 20 \rightarrow 10 \rightarrow 5 \rightarrow 16 \rightarrow
8 \rightarrow 4 \rightarrow 2 \rightarrow 1 = c_8$. We can therefore stop the
recomputation after just two steps, since $c_2=10$ has not changed.

Note that functorial tasks allow us to statically extract even more precise
information about their dependencies than applicative tasks. Indeed, we
statically know that a \hs{Task}~\hs{Functor} has exactly one \hs{dependency}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependency :: Task Functor k v -> k
dependency task = getConst $ run task Const
\end{minted}
\vspace{1mm}
